---
title: "Reproducible Research Course Project 2"
author: "Anjana Ramesh"
date: "7/28/2020"
output:
  html_document: 
    keep_md: yes
  pdf_document: default
---
# **Analysis of the Impact of Different Weather Conditons on Health and the Economy**

# Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

# Data

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site.

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

## **Assignment**

The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis. Your analysis can consist of tables, figures, or other summaries. You may use any R package you want to support your analysis.

**Questions**

Your data analysis must address the following questions:

  - Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
  - Across the United States, which types of events have the greatest economic consequences?
  
Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.

## **Setting Global Options**

```{r globalOptions}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, fig.width = 10, fig.height = 5,
                      fig.keep = 'all' ,fig.path = 'figures\ ', dev = 'png')
```

## **Loading and Examining Data**

```{r stormData}
# Loading packages
library(ggplot2)
library(ggthemes)
library(dplyr)
library(plyr)

# Reading the data
stormData <- read.csv(bzfile("repdata_data_StormData.csv.bz2"))

# Examining dimesions of stormData
dim(stormData)
```

```{r}
# Examining the variables of stormData
str(stormData)
```

## **Extracting Variables of Interest**

To analyse the impact of different weather conditions, it is  needful to extract the relevant variable data from the dataset.

- Weather Events Variable 
  - EVTYPE: type of weather event (TORNADO, FLOOD etc)

- Health variables
  - FATALITIES: number of deaths
  - INJURIES: number of injuries

- Economic variables 
  - PROPDMG: property damage
  - PROPDMGEXP: units for property damage value
  - CROPDMG: crop damage
  - CROPDMGEXP: units for crop damage value

```{r subsetStormData}
# Extracting the relevant variable data from stormData
variables <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
subsetStormData <- stormData[ , variables]

# Viewing the subset 
str(subsetStormData)

```

## **Transforming and Cleaning the Extracted Variables**

```{r eventType}

# Viewing the types that feature the most in the variable EVTYPE
sort(table(subsetStormData$EVTYPE), decreasing = TRUE)[1:10]

```

Grouping the types in the variable EVTYPE according to common keywords like HEAT, TORNADO, FLOOD etc-

```{r groupingTypes}
# Creating a new variable to transform EVTYPE into groups 
subsetStormData$EVENT <- "OTHER"

# Grouping types in EVTYPE by keywords

subsetStormData$EVENT[grepl("HEAT", subsetStormData$EVTYPE, ignore.case = TRUE)] <- "HEAT"
subsetStormData$EVENT[grepl("HAIL", subsetStormData$EVTYPE, ignore.case = TRUE)] <- "HAIL"
subsetStormData$EVENT[grepl("SNOW", subsetStormData$EVTYPE, ignore.case = TRUE)] <- "SNOW"
subsetStormData$EVENT[grepl("WIND", subsetStormData$EVTYPE, ignore.case = TRUE)] <- "WIND"
subsetStormData$EVENT[grepl("STORM", subsetStormData$EVTYPE, ignore.case = TRUE)] <- "STORM"
subsetStormData$EVENT[grepl("TORNADO", subsetStormData$EVTYPE, ignore.case = TRUE)] <- "TORNADO"
subsetStormData$EVENT[grepl("FLOOD", subsetStormData$EVTYPE, ignore.case = TRUE)] <- "FLOOD"
subsetStormData$EVENT[grepl("WINTER", subsetStormData$EVTYPE, ignore.case = TRUE)] <- "WINTER"
subsetStormData$EVENT[grepl("RAIN", subsetStormData$EVTYPE, ignore.case = TRUE)] <- "RAIN"

# Viewing the groups
sort(table(subsetStormData$EVENT), decreasing = TRUE)[1:10]

```


Crop Damage and Property Damage values are defined in exponential terms - 

```{r}
# Viewing Property Damage
table(subsetStormData$PROPDMGEXP)

```

```{r}
# Viewing Crop Damage
table(subsetStormData$CROPDMGEXP)

```

Creating factor variables to bring the values of damage to the same base (Millions of USD$) -

```{r propDmgValue}
# Creating new variable to standardise PROPDMGEXP to Millions USD
subsetStormData$propDmgValue <- 0.000001

# Setting values according to the exponential powers
subsetStormData$propDmgValue[subsetStormData$PROPDMGEXP =="1"] <- 0.00001
subsetStormData$propDmgValue[subsetStormData$PROPDMGEXP =="2"] <- 0.0001
subsetStormData$propDmgValue[subsetStormData$PROPDMGEXP =="3"] <- 0.001
subsetStormData$propDmgValue[subsetStormData$PROPDMGEXP =="4"] <- 0.01
subsetStormData$propDmgValue[subsetStormData$PROPDMGEXP =="5"] <- 0.1
subsetStormData$propDmgValue[subsetStormData$PROPDMGEXP =="6"] <- 1
subsetStormData$propDmgValue[subsetStormData$PROPDMGEXP =="7"] <- 10
subsetStormData$propDmgValue[subsetStormData$PROPDMGEXP =="8"] <- 100
subsetStormData$propDmgValue[subsetStormData$PROPDMGEXP =="B"] <- 1000
subsetStormData$propDmgValue[subsetStormData$PROPDMGEXP =="h"] <- 0.0001
subsetStormData$propDmgValue[subsetStormData$PROPDMGEXP =="H"] <- 0.0001
subsetStormData$propDmgValue[subsetStormData$PROPDMGEXP =="K"] <- 0.001
subsetStormData$propDmgValue[subsetStormData$PROPDMGEXP =="m"] <- 1
subsetStormData$propDmgValue[subsetStormData$PROPDMGEXP =="M"] <- 1

# Viewing property damage values
table(subsetStormData$propDmgValue)

```


```{r cropDmgValue}
# Creating a new variable to standardise CROPDMGEXP to Millions USD
subsetStormData$cropDmgValue <- 0.000001

# Setting values according to the exponential powers
subsetStormData$cropDmgValue[subsetStormData$CROPDMGEXP =="2"] <- 0.0001
subsetStormData$cropDmgValue[subsetStormData$CROPDMGEXP =="B"] <- 1000
subsetStormData$cropDmgValue[subsetStormData$CROPDMGEXP =="k"] <- 0.001
subsetStormData$cropDmgValue[subsetStormData$CROPDMGEXP =="K"] <- 0.001
subsetStormData$cropDmgValue[subsetStormData$CROPDMGEXP =="m"] <- 1
subsetStormData$cropDmgValue[subsetStormData$CROPDMGEXP =="M"] <- 1

# Viewing crop damage values
table(subsetStormData$cropDmgValue)

```

## **Aggregating Events for Health Variables**

```{r}

# Aggregating fatalities and injuries by type of event 
aggregateFatalitiesandInjuries <- ddply(subsetStormData, .(EVENT), summarise, 
                                       Total = sum(FATALITIES + INJURIES, na.rm = TRUE))
# Changing the col name
aggregateFatalitiesandInjuries$type <- "Fatalities and Injuries"

aggregateFatalitiesandInjuries
```

```{r}
# Aggregating fatalities by type of event 
aggregateFatalities <- ddply(subsetStormData, .(EVENT), summarise, 
                                       Total = sum(FATALITIES, na.rm = TRUE))
# Changing the col name
aggregateFatalities$type <- "Fatalities"

aggregateFatalities
```

```{r}
# Aggregating injuries by type of event 
aggregateInjuries <- ddply(subsetStormData, .(EVENT), summarise, 
                                      Total = sum(INJURIES, na.rm = TRUE))
# Changing the col name
aggregateInjuries$type <- "Injuries"

aggregateInjuries
```


```{r aggregateHealth}
# Combining all variables
aggregateHealth <- rbind(aggregateFatalities, aggregateInjuries)

# Separating Fatalities and Injuries by event
healthByEvent <- join(aggregateFatalities, aggregateInjuries, by = "EVENT", type = "inner")
healthByEvent

```

## **Aggregating Events for Economic Variables**

```{r}

# Aggregating property and crop damage by type of event 
aggregatePropandCropDmg <- ddply(subsetStormData, .(EVENT), summarise, 
                                       Total = sum(propDmgValue + cropDmgValue, na.rm = TRUE))
# Changing the col name
aggregatePropandCropDmg$type <- "Property and Crop Damage"

aggregatePropandCropDmg

```

```{r}

# Aggregating property damage by type of event 
aggregatePropDmg <- ddply(subsetStormData, .(EVENT), summarise, 
                                       Total = sum(propDmgValue, na.rm = TRUE))
# Changing the col name
aggregatePropDmg$type <- "Property Damage"

aggregatePropDmg

```

```{r}
# Aggregating crop damage by type of event 
aggregateCropDmg <- ddply(subsetStormData, .(EVENT), summarise, 
                                       Total = sum(cropDmgValue, na.rm = TRUE))
# Changing the col name
aggregateCropDmg$type <- "Crop Damage"

aggregateCropDmg
```

```{r aggregatEconomy}
# Combining all variables
aggregateEconomy <- rbind(aggregatePropDmg, aggregateCropDmg)

# Separating Fatalities and Injuries by event
economyByEvent <- join(aggregatePropDmg, aggregateCropDmg, by = "EVENT", type = "inner")
economyByEvent

```

# **Results**

# Question 1 - Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

```{r healthPlot}
# Transforming events in aggregateHeath to a factor
aggregateHealth$EVENT <- as.factor(aggregateHealth$EVENT)

# Plotting into ggplot2
healthPlot <- ggplot(data = aggregateHealth, aes(x = EVENT, y = Total, fill = EVENT)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  xlab("Events") +
  ylab("Fatalities and Injuries") +
  ggtitle("Most Harmful Events in the US with respect to Health") +
  theme_calc(base_family = "serif")

print(healthPlot)
```

# Question 2 - Across the United States, which types of events have the greatest economic consequences?

```{r economyPlot}
# Transforming events in aggregateEconomy to a factor
aggregateEconomy$EVENT <- as.factor(aggregateEconomy$EVENT)

# Plotting into ggplot2
economyPlot <- ggplot(data = aggregateEconomy, aes(x = EVENT, y = Total, fill = EVENT)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  xlab("Events") +
  ylab("Total in Millions of USD of Property and Crop Damage") +
  ggtitle("Most Harmful Events in the US with respect to Economic Consequences") +
  theme_calc(base_family = "serif") +
  scale_fill_calc()

print(economyPlot)
```

**---------------------------------------------------------------------------------------------------**